---
title: "Natural Gas Project"
author: "Fatemeh Khosravi & Dennis Myasnyankin"
date: "2022-12-1"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(knitr)
library(reticulate)
library(ggplot2)
library(tidyverse)
library(readxl)
library(fpp2)
library(zoo)
library(readr)
library(forecast)
library(dplyr)
library(caret)
library(tidyr)
library(imputeTS)
library(lubridate)
library(Metrics)
knitr::opts_chunk$set(echo = TRUE)
```

```{r global.options, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  fig.align = 'center',
  fig.path = 'figs/',
  echo = TRUE,
  message = FALSE,
  strip.white = TRUE,
  warning = FALSE
)
```

**Final Project**

```{r}

# Load in the data set and convert Data

gas.daily= read_csv("daily_csv.csv",
                      col_types = cols(Date = col_date(format = '%Y-%m-%d')))

# Set seed for reproducibility

set.seed(21)

# Explore initial data set

plot(gas.daily , ylab='Price (Dollars)', main='Time Plot of Daily Natural Gas Price', 
                    xlab='Time', type='l', ylim=c(0,20.1))
```

```{r}
# Check for missing values

missingValsCount <- sum(is.na(gas.daily))
missingValIndices <- unique (unlist (lapply (gas.daily, function (x) which (is.na (x)))))
cat("There is a total of ", missingValsCount, " missing value(s) in the data frame.")

# List missing values

print(gas.daily[missingValIndices,])
```

From the initial exploration above, we can observe that there is 1 missing value present in the data set. The column associated with the null value is "Price", which was unable to be gathered by the source for 01/05/2018. Since gaps within time series forecasting may lead to sub optimal performance, we will address this issue through imputation. 

```{r}
# Impute via forecast package

gas.daily.df <- gas.daily
gas.daily.df$Price <- na.interp(gas.daily.df$Price)

cat("The imputed price for 2018-01-05 is: ", 
    gas.daily.df[missingValIndices,]$Price, 
    "\nThe new working data set now has: ", 
    sum(is.na(gas.daily.df)), " missing value(s).")
```

```{r}
# Add month, year, weekday columns to df using lubridate package

gas.daily.df <- gas.daily.df %>%
  mutate(Month = month(gas.daily.df$Date)) %>% 
  mutate(Year = year(gas.daily.df$Date)) %>% 
  mutate(Week_Day = weekdays(gas.daily.df$Date))

#head(gas.daily.df, 1000)

```

```{r, fig.width=5,fig.height=7}
# Calculate the average price for each month

monthYearBreakdown <- gas.daily.df %>%
  group_by(Month, Year) %>%
  summarise(Monthly_Avg = mean(Price))


# Plot histograms of prices of Month

gas.daily.df %>%
  ggplot(aes(x=Price)) + geom_histogram(bins = 10) +
  facet_wrap(~ Year, ncol = 4) +
      labs(title = "Monthly Natural Gas Price Distribution by Year",
           y = "Price (USD)",
           x = "Month") + theme_bw(base_size = 10)
```




```{r}
# Aggregate daily price to monthly price

gas.monthly = gas.daily.df %>%
   mutate(Month = format(Date, '%Y-%m'))  %>%
   group_by(Month = format(Date , '%Y-%m')) %>%
   summarise(Price = round(mean(Price), 2))

# Drop last value

gas.monthly = gas.monthly[-c(285), ]   
```

```{r}
# Create daily and monthly time series

price.daily <- ts(gas.daily.df$Price, start = c(1997,1) , end = c(2020,8), frequency = 12)
price.monthly <- ts(gas.monthly$Price, start = c(1997,1) , end = c(2020,8), frequency = 12)

# Plot daily and monthly time series 

par(mfrow=c(2,1))
daily.plot <- plot(price.daily, ylab = "Price ( Dollars)",
     main = "Time Plot of Daily Natural Gas Prices")
monthly.plot <- plot(price.monthly, ylab = "Price ( Dollars)",
     main = "Time Plot of Monthly Natural Gas Prices")
```

```{r}
# Create moving average of monthly data for trend analysis

price.monthly.centered = ma(price.monthly, order = 12)

autoplot(price.monthly, series="Actual", 
        main=" Centered Moving Average with Window=12, Overlaid on Monthly Gas Price",
        ylab="Price (Dollars)")+
  autolayer(price.monthly.centered, series= "CenteredAvg")+
  theme_classic()
```

Plotting the centered moving average over the original data provides us with an idea of where certain points tend to lie within the time series, giving an idea of the level associated with the monthly natural gas price. However, it is hard to understand the trend and seasonality still needs to be examined.


```{r}
# Differencing 

detrended.ts <- diff(price.monthly, lag = 1)
deseasoned.ts <- diff(price.monthly, lag = 12)
twice.differenced.ts <- diff(diff(price.monthly, lag=12), lag=1)

# Plots for systematic component analysis: level, trend, seasonality

layout(matrix(c(1,1,2,3,4,5), 3, 2, byrow = TRUE))

acf.plt <- Acf(price.monthly , lag.max = 240, 
               main = "Autocorrelation Plot of Monthly Gas Prices")
original.plt <- plot(price.monthly, ylab = "Price (Dollars)", 
                     main = "Original Monthly Gas TS")
detrended.plt <- plot(detrended.ts, ylab = "Price (Dollars)", 
                      main = "De-Trended Monthly Gas TS")
deseasoned.plt <- plot(deseasoned.ts, ylab = "Price (Dollars)", 
                       main = "De-Seasoned Monthly Gas TS")
twice.differenced.plt <- plot(twice.differenced.ts, ylab = "Price (Dollars)",
                              main = "Twice Differenced Monthly Gas TS")
```

The plots above depict the Autocorrelation Function...

```{r}
# Create benchmark model with forecast package's naive() function

fixed.nValid <- 20
fixed.nTrain <- length(price.monthly) - fixed.nValid
train.ts <- window(price.monthly, start = c(1997, 1), end = c(1997, fixed.nTrain))
valid.ts <- window(price.monthly, start = c(1997, fixed.nTrain + 1), 
                   end = c(1997, fixed.nTrain + fixed.nValid))
naive.model <- naive(train.ts, h = fixed.nValid)

print(naive.model)
```


```{r message=FALSE, warning=FALSE}

# Apply exponential smoothing using ets()

  # ZZZ -> Auto-selected Errors, Auto-selected Trend, Auto-selected Seasonality

model.ets <- ets(train.ts, model = "ZZZ", restrict = FALSE, 
                 allow.multiplicative.trend = TRUE)

summary(model.ets)

# Generate forecasts

ets.pred = forecast(model.ets, h=fixed.nValid)

# Get RMSE on the validation period

ets.rmse <- rmse(valid.ts, ets.pred$mean)
cat("\nETS Validation Set RMSE: ", ets.rmse)

# Create Plot

autoplot(train.ts, series = "Training", ylab = 'Price (Dollars)', xlab = 'Time', 
         bty = "l", xaxt = "n", flty= 2,
         main = "Forecasted Monthly Gas Price using Exponential Smoothing") +
  autolayer(valid.ts, series = "Actual") +
  autolayer(ets.pred, series = "ETS Forecast", alpha = 0.3) +
  coord_cartesian(xlim = c(2015, 2021))
```

The exponential smoothing function returns a model with the following ETS parameters (M,N,N). Suggesting multiplicative error, no trend, and no seasonality within the data set. This could mean we should look into a smaller portion of the time series (for instance, spanning 10 years instead of 20) in order for trend and seasonality to be applicable.


```{r}
# Model Linear Regression

model.lm = tslm(train.ts ~ trend + I(trend ^2)+I(trend ^3)+season)
summary(model.lm)

# Generate forecasts

lm.pred = forecast(model.lm, h = fixed.nValid)

# Get RMSE on the validation period

lm.rmse <- rmse(valid.ts, lm.pred$mean)
cat("\nLinear Regression Validation Set RMSE: ", lm.rmse)

# Create Plot

autoplot(train.ts, series = "Training", ylab = 'Price (Dollars)', xlab = 'Time', 
         bty = "l", xaxt = "n", flty= 2,
         main = "Forecasted Monthly Gas Price using Linear Regression") +
  autolayer(valid.ts, series = "Actual") +
  autolayer(lm.pred, series = "Linear Regression Forecast", alpha = 0.3) +
  coord_cartesian(xlim = c(2015, 2021))

```


```{r}
# Model ARIMA Automatically

model.auto = auto.arima(train.ts)
summary(model.auto)

# Generate forecasts

auto.pred = forecast(model.auto, h = fixed.nValid)

# Get RMSE on the validation period

auto.rmse <- rmse(valid.ts, auto.pred$mean)
cat("\nAuto Arima Validation Set RMSE: ", auto.rmse)

# Create Plot

autoplot(train.ts, series = "Training", ylab = 'Price (Dollars)', xlab = 'Time', 
         bty = "l", xaxt = "n", flty= 2,
         main = "Forecasted Monthly Gas Price using Auto ARIMA") +
  autolayer(valid.ts, series = "Actual") +
  autolayer(auto.pred, series = "Auto ARIMA Forecast", alpha = 0.4) +
  coord_cartesian(xlim = c(2015, 2021))
```

```{r}
# Model ARIMA Manually

model.arima = arima(train.ts, order=c(6,2,2), seasonal = c(2,1,1))
summary(model.arima)

# Generate forecasts

arima.pred = forecast(model.arima, h = fixed.nValid)

# Get RMSE on the validation period

arima.rmse <- rmse(valid.ts, arima.pred$mean)
cat("\nManual Arima Validation Set RMSE: ", arima.rmse)

# Create Plot

autoplot(train.ts, series = "Training", ylab = 'Price (Dollars)', xlab = 'Time', 
         bty = "l", xaxt = "n", flty= 2,
         main = "Forecasted Monthly Gas Price using Manual ARIMA") +
  autolayer(valid.ts, series = "Actual") +
  autolayer(arima.pred, series = "Manual ARIMA Forecast", alpha = 0.4) +
  coord_cartesian(xlim = c(2015, 2021))
```

```{r}
# Model Neural Nets

model.nnet = nnetar(train.ts, p = 6, P = 2, size = 5)
summary(model.nnet)

# Generate forecasts

nnet.pred = forecast(model.nnet, h = fixed.nValid)

# Get RMSE on the validation period

nnet.rmse <- rmse(valid.ts, nnet.pred$mean)
cat("\nNeural Net Validation Set RMSE: ", nnet.rmse)

# Create Plot

autoplot(train.ts, series = "Training", ylab = 'Price (Dollars)', xlab = 'Time', 
         bty = "l", xaxt = "n", flty= 2,
         main = "Forecasted Monthly Gas Price using Neural Networks") +
  autolayer(valid.ts, series = "Actual") +
  autolayer(nnet.pred, series = "Neural Net Forecast", alpha = 0.4) +
  coord_cartesian(xlim = c(2015, 2021))
```




```{r}
#no seasonal difference
pacf(train.ts, lag.max = 60, main=" No Differencing")
acf(train.ts, lag.max = 144,main=" No Differencing")

#one seasonal difference 
pacf(diff(train.ts, lag=12), lag.max = 60,main=" One Seasonal Difference")
acf(diff(train.ts, lag=12), lag.max = 60,main=" One Seasonal Difference")

# two regular difference
pacf(diff(diff(diff(train.ts, lag=12))), lag.max = 60,
     main=" One Seasonal Difference and Two Regular Difference")
acf(diff(diff(diff(train.ts, lag=12))), lag.max = 60,
     main=" One Seasonal Difference and Two Regular Difference")
```
